---
title: "Gas Prices in Brazil"
output:
  pdf_document: default
  html_notebook: default
---

In this project, my intent is to show if there is a historically difference in the mean price of a given source of energy among states within a given region in Brazil. Basically, my intent is to practice the use of ANOVA and Tukey tests.

That said, I'll stick to using only the Mean Market Value price as a target variable. I'll also study data based on final dates that are greater than a given start date. I won't study the data on a given date (for example, the last one) due to the very few ammount of records I'd have. Thus the need to use some historical data as well.


```{r}
library('tidyr')
library('dplyr')
library('readr')
```


```{r}
gasPrices_raw <- read.table(file = '2004-2019.tsv', sep = '\t', header = TRUE, encoding="UTF-8")
str(gasPrices_raw)
```



```{r}
head(gasPrices_raw)
#print.data.frame(head(gasPrices_raw))
```

I'll create a new dataframe to change some columns and factor names and to parse the final date column from factor to date format.


```{r}
gasPrices_temp <- gasPrices_raw

names(gasPrices_temp)[3] <- 'finalDate'
names(gasPrices_temp)[4] <- 'region'
names(gasPrices_temp)[5] <- 'state'
names(gasPrices_temp)[6] <- 'product'
names(gasPrices_temp)[9] <- 'meanMarketValue'

levels(gasPrices_temp$region) <- c('Center-West', 'Northeast',
                                   'North', 'Southeast', 'South')
levels(gasPrices_temp$product)[5] <- 'OLEO DIESEL'
levels(gasPrices_temp$product)[6] <- 'OLEO DIESEL S10'

gasPrices_temp$finalDate <- as.Date(gasPrices_temp$finalDate)
```


I'll define the start date mentioned earlier. Also, I'll define a giver product to study, since the study of all products would result in a long publication.


```{r}
startDate <- as.Date('01/08/2018', '%d/%m/%Y')
productOfInterest <- 'GASOLINA COMUM'

gasPrices_allProducts <- gasPrices_temp %>% 
                            filter(finalDate>startDate) %>% 
                            select(finalDate, region, state,  product, meanMarketValue) %>% 
                            arrange(state) %>%
                            arrange(region)

gasPrices <- gasPrices_allProducts %>% 
                filter(product==productOfInterest) 

head(gasPrices)
```


```{r}
library(ggplot2)
library(gridExtra)

evolutionPerRegion <- gasPrices %>% group_by(finalDate, region) %>% summarize(mean = mean(meanMarketValue))

meanPerState <- gasPrices %>% group_by(finalDate, region) %>% summarize(mean = mean(meanMarketValue)) %>%
  ungroup() %>%
  select(region, mean)

```


```{r, fig.width=9, fig.height=12}
p1 <- ggplot(aes(x=finalDate, y=mean, color=region), data=evolutionPerRegion) +
  geom_line() + 
  labs(color='Region: ') +
  xlab('Final Date') + 
  ylab('Mean Market Value') + 
  ggtitle('Evolution of fuel price amongs regions') + 
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "bottom",
        panel.background = element_rect(fill = 'white'),
        panel.grid.major = element_line(alpha('black', 0.4), linetype='dashed', size=.1),
        panel.grid.minor = element_line(alpha('black', 0.2), linetype='dashed', size=.1)) +
  
  guides(fill=guide_legend(title="New Legend Title"))

p2 <- ggplot(aes(x=region, y=mean, color=region), data=meanPerState) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) + 
  scale_x_discrete() + 
  ylab('Mean Market Value') + 
  ggtitle('Evolution of fuel price in North region') + 
  xlab('') + 
  ylab('') + 
  theme(plot.title = element_text(hjust = 0.5)) + 
  theme(legend.position = "none",
        panel.background = element_rect(fill = "white"))
  #      axis.title.x=element_blank())
  #      axis.text.x=element_blank(),
  #      axis.ticks.x=element_blank())

margin <- theme(plot.margin = unit(c(1,1,1.5,1), "cm"))
grid.arrange(grobs=lapply(list(p1, p2), "+", margin))


```





```{r}
for (region_str in unique(gasPrices$region)){
  anova <- aov(meanMarketValue ~ state, gasPrices %>% filter(region==region_str))
  print('***************************************')
  print(region_str)
  print('***************************************')
  print(summary(anova))
  print(TukeyHSD(anova, ordered=TRUE))
  cat(rep('\n', 2))
}
```



```{r}
meanByState <- gasPrices %>% 
                   group_by(region, state) %>% 
                   summarize(mean = mean(meanMarketValue)) %>%
                   top_n(2, mean)

print.data.frame(meanByState)
```



###############################


```{r}
for (prod in unique(gasPrices_allProducts$product)){
topMeanValues <- gasPrices_allProducts %>% 
                  filter(product==prod) %>%
                  select(state, meanMarketValue) %>%
                  group_by(state) %>%
                  summarize(mean = mean(meanMarketValue)) %>%
                  arrange(desc(mean))
cat(c('product: ', prod, '\n'))
print.data.frame(topMeanValues[1:5,])
cat('\n')
}

```


















